### 2019-12-17 manuscript/main.tex
more work on methods and discussion

Clinicians are generally interested in inferring transmission links between HIV-infected individuals to improve HIV treatment and prevention. In the absence of reliable patient contact histories, phylogenetic analysis of viral sequence data can be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share a transmission link in the real, unobservable transmission network. The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was recently identified as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.

While historically the phylogenetic inference power of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic and transmission cluster analyses, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing. In a recent meta study, Novitsky \emph{et al.} (forthcoming) found that 102 out of 107 published studies of HIV cluster inference between 2016 and 2019 used the \emph{pol} region alone.

@@

Most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny. Accordingly, researchers often summarize the within-host variation present in NGS data sets with a consensus sequence. In the larger context of phylogenetic methods (e.g. beyond their application to HIV sequences), this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}. In the context of HIV, tihs assumption is that the consensus sequence adequately captures most of the relevant information about the distribution of sequences that might be constructed from the within-host variation given the NGS data set. Some studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.

In this study, we examine whether this assumption underlying the consensus approach holds on a data set comprising all newly HIV-diagnosed individual in the first half of 2013 from the largest HIV center in Rhode Island, USA. We present a new analytical method called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of phylogenetic analysis and cluster inference when using the consensus approach.

@@

Our study and data collection were approved by Lifespan's Institutional Review Board. We collected viral sequences from patients newly diagnosed with HIV during 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA. Inclusion criteria were: (i) HIV-infected adults, 18 years of age or older; (ii) diagnosed with HIV during 2013; and (iii) available \emph{pol} sequence from routine drug resistance testing. We obtained whole genome viral sequences for the 37 included individuals using both Sanger and NGS. Patient identifiers were removed and all analysis was conducted with de-identified sequence data.

We introduce a new approach for incorporating within-host variation into phylogentic analysis, called \emph{profile sampling}. We start by aligning each individual's NGS reads to all reference database of all sequences from the Los Alamos National Lab HIV Database with the probabilistic multiple sequence aligner HMMER version 3.1b2 \parencite{eddy}, using a variant of the \emph{hivmmer} pipeline \parencite{howison} that has been extended to support near-whole genome HIV data. Next, we convert the resulting multiple sequence alignment to a profile hidden Markov model (pHMM). Because pHMMs are probabilistic representations of alignments, we can sample genomic sequences from each individual's pHMM.

sequences further allows us to estimate the probability of a molecular transmission	228
cluster given the intra-patient variation.  To do so, we first infer a multiple sequence	229
alignments and phylogenetic tree for each sampled genomic sequence. Then, we	230
compute the bootstrap support [37] for each sampled tree and use a 99% bootstrap	231
cutoff to identify transmission clusters.  Phylogenetic support value thresholds are the	232
most common approach used to identify transmission clusters [9], hence our decision to	233
apply it to our workflow.  The number of times a particular transmission cluster is	234
identified in the set of phylogenetic tree samples divided by the number of samples is	235
taken as the probability of that transmission cluster given the  intra-patient variation.	236
For  the profile-sampling approach we use the well-established pHMM  alignment tool	243
HMMER 3.1b2 [38] to build, align and sample from the pHMMs.

For both approaches we estimate multiple alignments across individuals using mafft version 7.305b [39] and maximum-likelihood phylogenies with bootstrap replicates using RAxML version 8.2.10.

@@

\subsection*{Profile sampling captures inter-host diversity}

@@

The true HIV transmission network is often unknown, but phylogenetic trees of patient viral sequences can provide information on transmission clusters that identify links between patients [1] as well as relevant features of transmission dynamics. [22, 23]. Current phylogenetic approaches to transmission inference utilize a single summary consensus sequence per patient, which ignores the extensive intra-patient viral diversity [6, 18].

>>>>

Clinicians are generally interested in inferring transmission links between HIV-infected individuals to improve HIV treatment and prevention. In the absence of reliable patient contact histories, phylogenetic analysis of viral sequence data can be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share a transmission link in the real, unobservable transmission network. The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was delineated as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.

While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing. In a recent meta study, Novitsky \emph{et al.} (forthcoming) found that 102 out of 107 published studies of HIV cluster inference between 2016 and 2019 used the \emph{pol} region alone.

@@

Most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny. Accordingly, researchers often summarize the within-host variation present in NGS data sets with a consensus sequence. In the larger context of phylogenetic methods (e.g. beyond their application to HIV sequences), this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}. In the context of HIV, tihs assumption is that the consensus sequence adequately captures the relevant information about the distribution of sequences that might be constructed given the within-host variation measured in deep sequencing through NGS. Some previous studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.

In this study, we examine how well this assumption underlying the consensus approach holds on a data set comprising all newly HIV-diagnosed individual in the first six months of 2013 from the largest HIV center in Rhode Island, USA. We present a new analysis method called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of the consensus approach for phylogenetic analysis and cluster inference.

@@

\subsection*{Data collection and sequencing}

Our study and data collection were approved by Lifespan's Institutional Review Board. We collected viral sequences from 37 individuals newly diagnosed with HIV during 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA. Inclusion criteria were: (i) HIV-infected adults, 18 years of age or older; (ii) diagnosed with HIV during the first six months of 2013; and (iii) available \emph{pol} sequence from routine drug resistance testing. Patient identifiers were removed and all analysis was conducted with de-identified sequence data.

We obtained near-whole genome viral sequences for the 37 participants using both Sanger and NGS sequencing methods. Blood specimens were obtained from participants with their consent and processed to isolate peripheral blood mononuclear cells (PMBC), buffy coats, and plasma. From the collected plasma, total nucleic acid was extracted for genotyping. An in-house genotyping assay was used to generate the near-whole genome based on previously published methods \parencite{nadai, di_giallonardo}. For each sample, two cDNA templates were generated by SuperscriptIII First Strand Synthesis System (Thermofisher, Carlsbad, CA), followed by eight separate nested PCR reactions; these eight amplicons span the near-whole genome of HIV.  Final amplicon products were sequenced by the Sanger method using 3100 Genetic Analyzer (Applied Biosystems, Foster City, CA) and sequenced by NGS using Nextera XT DNA Library Prep chemistry (Illumina, San Diego, CA) to generate multiplexed libraries for Illumina's MiSeq platform with 250 base paired-end reads. Sanger consensus sequences were generated manually using Sequencher version 5.2.4 (Gene Codes, Ann Arbor, MI) to confirm degenerate nucleotides. NGS data were processed and demultiplexed using BaseSpace cloud based application? (Illumina, San Diego, CA).

\subsection*{Profile sampling}

We introduce a new approach for incorporating within-host variation into phylogentic analysis, called \emph{profile sampling}. We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (pending release as hivmmer version 0.3.0). Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to all group M reference sequences from the Los Alamos National Lab HIV Database with the probabilistic multiple sequence aligner HMMER version 3.1b2 \parencite{eddy}, and produces a codon frequency table across the near-whole HIV genome, which we refer to as the individual's HIV \emph{profile}.

We construct fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the profile. The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level. We sample 500 sequences from each individual's profile, then collate the sampled sequences into 500 profile-sampled data sets, each having one sampled sequence per individual. We perform phylogenetic inference on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}. In addition to the 500 profile-sampled phylogenies, we infer two additional phylogenies from the NGS consensus sequences and Sanger consensus sequences. We perform cluster inference on these phylogenies using Cluster Picker \parencite{ragonnet-cronin} with thresholds of 80\% bootstrap support and 4.5\% genetic distance. Analysis source code is available from \url{https://github.com/kantorlab/hiv-profile-sampling}.

In addition to performing these analyses on the near-whole genome sequences (``wgs''), we also perform them on subsets of the sequences in three clinicially relevant regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene. The prrt region and \emph{int} are routinely sequenced in clinical care to detect drug resistance mutations and inform clinical choices of anti-retroviral therapy. The \emph{env} region is... Rami, why is this routinely sequenced??

@@

\subsection*{Profile sampling captures within-host diversity}

We examine how much

@@

A limitation of our study is the small number of participants. The participants do have a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, who met the inclusion criteria. However, the overall size of the HIV epidemic in Rhode Island is much larger, estimated as 2,396 individuals in 2016 \parencite{ridoh}. In future work, we hope to apply the profile sampling method to larger NGS data sets, to assess if cluster inference from Sanger versus NGS data agrees or differs especially for clusters larger than three individuals, which was the largest cluster size in our current data set.

Our construction of HIV profiles from NGS data is limited by the accuracy of the NGS assays themselves. The codon frequencies in the profiles may be biased measures of the true within-host diversity because of biases in PCR amplification during sample preparation. Sequencing protocols such as Primer ID \cite{jabara} have been introduced to reduce and correct for these biases. Unfortunately, the Primer ID protocol supports only a limited region of the HIV genome, and has not yet been extended to support near-whole genome sequencing of HIV, but future work could study this limited region to determine if cluster inference changes with the implementation of a Primer ID protocol.

The true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \cite{fauci}. Current phylogenetic approaches do not fully utilize the information on within-host diversity available from deeper and longer NGS sequencing technologies. As NGS data sets are increasingly available and more fully sample from the current HIV epidemic,
